Volunteers in the Clouds: an Architecture for Low Cost and Potentially Massive Distributed Evolutionary Computation

### 2015-11-17 special-issue-ci/Response to reviewers.tex

4. Style: Paper is generally written in a clear way. Statements like
"But fair is fair," (page 6), sound a bit colloquially in a scientific
paper and may be omitted. However, my main concern is the readability
of the graphs. General remark - font used for description of axes

\end{quote}

We have eliminated colloquialisms from the text.

\begin{quote}
should be significantly larger. In Fig. 2 the authors can change scale
for axe Y, as the box for the population of 1024 individuals is very
small.  What do the authors mean by "times"? Time in seconds? I am not
sure if this Figure is necessary at all as the description is
understandable without it. In Fig.4 all graphs are similar and only
the one in the middle is analyzed, so maybe others are not necessary?
\end{quote}

The labels, and indeed the label size, are not too important; the
graph shape is. That is one reason they are grayed out. However, we
have changed as instructed Figure 2, which is more readable now. We do
mean times to solution, shortened for clarity. We have also clarified
that on the caption. We also think that all graphs are needed in Fig
4, since they represent different experiments, and we have added a
sentence to highlight that the {\em tail} appears in all three.

\begin{quote}
The same is with Fig. 5. What does "x" stand for in the description of
axe X? (the same is with Fig. 8).
\end{quote}

$x$ has been changed to {\sf rank}, which is the usual way of
representing Zipf's law, a measure (in this case number of {\tt PUT}s
vs. rank. We have also changed size, color and sigils to represent
each point to make it a bit clearer. 
%Still fig 8 to go

### 2015-11-17 special-issue-ci/Response to reviewers.tex

\begin{quote}
1. Contribution: The paper presents the idea of a web browser based distributed evolutionary computing.
the system called NodIO consists of a REST server (coded in Node.js) and clients performing
evolutionary algorithm coded in JavaScript and embedded in a web page. Although the idea is not
entirely new, the authors show how Web Workers (introduced in HTML5) and the mechanism of
automatic restart of computations on the client's side can improve the performance of the system. The
state of the art is comprehensive and provides good introduction into the subject.

2. Relevance: In my opinion the system described in the paper is relevant to "Computational Intelligence
Software" special issue. Thanks to both development of Internet technology and continuous increase in
computing power of electronic devices, including mobile devices, the web browser based distributed
computing can be a cheap alternative for grid and high-performance computing systems that require
high capital expenditures and consume a lot of energy. Evolutionary computing system based on the
architecture presented inthe paper can be potentially used for solving many different time consuming
problems, however, the authors limited their experiments to a simple "Trap" function that can be easily
solved on a desktop machine, what may not be convincing for a reader looking for an effective
optimization tool.
\end{quote}

It is true that that function can be easily solved with any desktop
machine, but we can still do that {\em and} use a volunteer
computation framework, which will add performance to the desktop
machine. Our intention was more to prove what kind of performance can
be expected and how this performance is distributed, although we have
also proved in the last section how the performance achieved by a
volunteer computer system can, in fact, be better than a desktop, even
a powerful one. This has also been clarified in the introduction and
conclusions.

At any rate, we have also performed a new experiment with the
Rastrigin function, which in fact cannot be solved in a few minutes
(or seconds, as in the case of Trap) by a desktop system.

### 2015-11-17 special-issue-ci/Response to reviewers.tex

$x$ has been changed to {\sf rank}, which is the usual way of
representing Zipf's law, a measure (in this case number of {\tt PUT}s
vs. rank. We have also changed size, color and sigils to represent
each point to make it a bit clearer.

\begin{quote}

5. Length: As I mentioned above the authors should think of removing some of the unnecessary graphs
in order to improve readability. They admit themselves that it is hard to conclude from some of the
graphs (Fig. 5 and Fig. 8). Nevertheless, overall length of the paper is fine.

Finally, since the reviewer thinks that length is fine and for the
reasons outlined above, we have decided to keep the graphs and eplain
them a bit better.

\begin{quote}
6. Abstract: Abstract is adequate, moreover it puts an accent on the analysis of volunteers' behavior in a
distributed computing system, contrary to the description in the main text of the paper. However, it
lacks the information about the second version of the system called NodIO-W2 based on the Web
Worker architecture.
\end{quote}

we have extended the abstract stressing our main points, that were
first to test the validity and measure the performance of our
framework NodIO and then to use the measures as first steps to a model
of volunteer uses.

\begin{quote}
Comments to Authors:

### 2015-11-17 special-issue-ci/volunteer-clouds-ieee-ci-si.tex
Changing figures
Refs #48
Modified plot for fig 8
And a bit of the biblio. Closes #48
Extends abstract
To stress the main points and closes #49

\begin{abstract}
Volunteer computing is used routinely in science and industry, but it
still has a long way to go if it is going to become a mainstream technology in
the same way the cloud technologies in general have. The lack of general
purpose and open source tools
with analytics built in could be one of the reasons why this is so, as
well as the lack of proofs of concept. That is why in this paper we
present {\sf NodIO}, a framework used to create client-server architectures for distributed
evolutionary algorithms, written entirely in JavaScript, which needs
minimal server support to work and is able to support persistent,
asynchronous, distributed evolutionary algorithms running in the
browser.  The architecture, along with some initial results on the
expected performance and scaling
behavior is presented. We also use the algorithm data to
find out what kind of patterns the user behavior follow and try to fit
them to some statistical distribution.
\end{abstract}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of {\tt PUT}s per unique IP and fit to a Generalized
  Extreme Value distribution (in red). From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows a S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, and of the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it drops sharply implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in the
distribution of HTTP {\tt PUT}s, equivalent to the number of
generations divided by 100, contributed by every user, which is shown
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of {\tt PUT}s per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, negative towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{ips-time-ww.png}
\caption{Time employed in every experiment vs. number of IPs (in
  abscissas). The red dot and line show the average and standard
  deviation, the shade of blue the number of experiments with the same
  number of IPs. } 
\label{fig:ipstime:w2}
\end{figure}
%
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value (GEV) and Weibull distribution of
  the number of {\tt PUT}s per worker. \label{tab:puts:ww}}
\begin{center}
\begin{tabular}{cccc}
\hline
Distribution & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
GEV & 18.275 $\pm$ 0.83719  &  25.962  $\pm$ 1.22015 & 1.242   $\pm$
0.04746 \\
Weibull & ND & 75.11457519 $\pm$ 2.97470251  & 0.69714410 $\pm$ 0.01383976 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.49\linewidth]{gev-fit-ww.png}
\includegraphics[width=0.49\linewidth]{weibull-fit-ww.png}
\caption{Ranked number of {\tt PUT}s per worker and GEV fit (in red, right) and Weibull (in red, left) for the {\sf NodIO$^2$} model.}  
\label{fig:gev:w2}
\end{figure}
%
We are also interested in measuring the scaling properties in this
model, that is, the relationship between the number of IPs
participating in an experiment and the time it takes to complete
it. This is shown in Figure \ref{fig:ipstime:w2}, which shows every
experiment in a time (in milliseconds) vs. number of workers participating
in it. Although there is not a clear trend, the graph seems to say
that the time needed for finding the solution in this evolutionary
algorithm does not depend on the number of workers participating in
it. This does not mean that it is independent on the number of {\em
  simultaneous} users, which is probably the case although it is more
difficult to measure. There seems to be a trend towards a decreasing
standard deviation in the time, with more volunteers adding {\em
  robustness} to an experiment and certainty in the time it is taking;
however, more experiments would be needed to check this hypothesis.

>>>>

\begin{abstract}
Volunteer computing is used routinely in science and industry, but it
still has a long way to go if it is going to become a mainstream
technology in the same way the cloud technologies in general have. The
lack of general purpose and open source tools with analytics built in
could be one of the reasons why this is so, as well as the lack of
proofs of concept. That is why in this paper we present {\sf NodIO}, a
framework used to create client-server architectures for distributed
evolutionary algorithms, written entirely in JavaScript, which needs
minimal server support to work and is able to support persistent,
asynchronous, distributed evolutionary algorithms running in the
browser, as well as any other client consuming its REST application
programming interface.  The architecture, along with some initial
results on the expected performance and scaling behavior when the
number of users is increased is presented. We also use
the algorithm data to find out what kind of patterns the user behavior
follow by examining the resources they devote to our experiment, and
try to fit them to some statistical distribution to create the bases
of a model of user behavior in volunteer computing experiments of this
kind that can be used by researchers to predict its performance. 
In general, our experiments prove first that NodIO is a flexible and
low-overhead platform to perform distributed computation experiments
that include volunteers and that user behaviour follows some patterns,
and, thus, are amenable to a model.
\end{abstract}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of {\tt PUT}s per unique IP and fit to a Generalized
  Extreme Value distribution (in red). From left to right, experiments
  4/4, 4/24 and 7/31.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows a S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, and of the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it drops sharply implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in the
distribution of HTTP {\tt PUT}s, equivalent to the number of
generations divided by 100, contributed by every user, which is shown
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of {\tt PUT}s per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, negative towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{ips-time-ww.png}
\caption{Time employed in every experiment vs. number of IPs (in
  abscissas). The red dot and line show the average and standard
  deviation, the shade of blue the number of experiments with the same
  number of IPs. } 
\label{fig:ipstime:w2}
\end{figure}
%
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value (GEV) and Weibull distribution of
  the number of {\tt PUT}s per worker. \label{tab:puts:ww}}
\begin{center}
\begin{tabular}{cccc}
\hline
Distribution & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
GEV & 18.275 $\pm$ 0.83719  &  25.962  $\pm$ 1.22015 & 1.242   $\pm$
0.04746 \\
Weibull & ND & 75.11457519 $\pm$ 2.97470251  & 0.69714410 $\pm$ 0.01383976 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.49\linewidth]{gev-fit-ww.png}
\includegraphics[width=0.49\linewidth]{weibull-fit-ww.png}
\caption{Ranked number of {\tt PUT}s per worker and GEV fit (in red, right) and Weibull (in red, left) for the {\sf NodIO$^2$} model.}  
\label{fig:gev:w2}
% File ../data/ips-puts.R
\end{figure}
%
We are also interested in measuring the scaling properties in this
model, that is, the relationship between the number of IPs
participating in an experiment and the time it takes to complete
it. This is shown in Figure \ref{fig:ipstime:w2}, which shows every
experiment in a time (in milliseconds) vs. number of workers participating
in it. Although there is not a clear trend, the graph seems to say
that the time needed for finding the solution in this evolutionary
algorithm does not depend on the number of workers participating in
it. This does not mean that it is independent on the number of {\em
  simultaneous} users, which is probably the case although it is more
difficult to measure. There seems to be a trend towards a decreasing
standard deviation in the time, with more volunteers adding {\em
  robustness} to an experiment and certainty in the time it is taking;
however, more experiments would be needed to check this hypothesis.
